---
title: "Day 1 Part 2 R Exercise Julianne Murphy"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load the data

Load the `gapminder` dataset.
```{r chunk name}
gapminder <- read.csv(here::here("data/gapminder5.csv"), stringsAsFactors=FALSE)
```

# If Statement

## Asssessing the inclusion of records in 2002 and 2012

Use an if() statement to print a suitable message reporting whether there are any records from 2002 in the gapminder dataset. 

```{r}
year<-2002
if(any(gapminder$year == year)){
   print(paste("Record(s) for the year",year,"found."))
} else {
  print(paste("No records for year",year))
}
```

Now do the same for 2012.
```{r}
year<-2012
if(any(gapminder$year == year)){
   print(paste("Record(s) for the year",year,"found."))
} else {
  print(paste("No records for year",year))
}
```

# Loop and If Statements

## Mean life expectancy by country for countries whose population is below the mean for the dataset

```{r}
overall_mean <- mean(gapminder$pop)

for (i in unique(gapminder$country)) {
   country_mean <- mean(gapminder$pop[gapminder$country==i])
   
   if (country_mean < overall_mean) {
     mean_le <- mean(gapminder$lifeExp[gapminder$country==i])
     print(paste("Mean Life Expectancy in", i, "is", mean_le))
   } 
} # end for loop
```

## Mean life expectancy by continent if it is smaller than 50, between 50 and 70, or greater than 70.

```{r}
lower_threshold <- 50
upper_threshold <- 70
 
for (i in unique(gapminder$continent)){
    tmp <- mean(gapminder$lifeExp[gapminder$continent==i])
    
    if (tmp < lower_threshold){
        print(paste("Average Life Expectancy in", i, "is less than", lower_threshold))
    }
    else if (tmp > lower_threshold & tmp < upper_threshold){
        print(paste("Average Life Expectancy in", i, "is between", lower_threshold, "and", upper_threshold))
    }
    else {
        print(paste("Average Life Expectancy in", i, "is greater than", upper_threshold))
    }
   
}
```

# Exercise: Write Functions 

## Column names and data type
Create a function that given a data frame will print the name of each column and the class of data it contains. Use the gapminder dataset.  Hint: Use `mode()` or `class()` to get the class of the data in each column. Remember that `names()` or `colnames()` returns the name of the columns in a dataset.

Note: Some of these were taken or modified from https://www.r-bloggers.com/functions-exercises/

```{r}
data_frame_info <- function(df) {
  cols <- names(df)
  for (i in cols) {
    print(paste0(i, ": ", mode(df[, i])))
  }
}
data_frame_info(gapminder)
```

## Vector mean and standard deviation
Create a function that given a vector will print the mean and the standard deviation of a **vector**, it will optionally also print the median. Hint: include an argument that takes a boolean (`TRUE`/`FALSE`) operator and then include an `if` statement.

```{r}
vector_info <- function(x, include_median=FALSE) {
  print(paste("Mean:", mean(x)))
  print(paste("Standard Deviation:", sd(x)))
  if (include_median) {
    print(paste("Median:", median(x)))
  }
}

le <- gapminder$lifeExp
vector_info(le, include_median = F)
vector_info(le, include_median = T)
```

# Analyzing the Relatioship 

## GDP per capita and life expectancy over time
What is the relationship between GDP per capita and life expectancy? Does this relationship change over time? (Hint: Use the natural log of both variables.)
```{r}
gapminder$vec_log_gdpPercap <- log(gapminder$gdpPercap)
gapminder$log_lifeExp <- log(gapminder$lifeExp)
cor(gapminder$log_lifeExp, gapminder$vec_log_gdpPercap)
```

Both GDP per capita and life expectancy increase over time 
```{r}
for (j in years) {
    mean_le <- mean(gapminder$log_lifeExp[gapminder$continent==i & gapminder$year == j], na.rm=T)
    mean_gdp <- mean(gapminder$vec_log_gdpPercap[gapminder$continent==i & gapminder$year==j], na.rm=T)
  print(paste0(j, " -> Mean Life Expectancy: ", mean_le, "; Mean GDP: ", mean_gdp))
  }
```

### Plots
```{r}
smoothScatter(gapminder$log_lifeExp, gapminder$vec_log_gdpPercap)
plot(gapminder$log_lifeExp, gapminder$vec_log_gdpPercap)
```


## GDP per capita and life expectancy by continent
Does the relationship between GDP per capita and life expectancy vary by continent? Make sure you divide the Americas into North and South America.

Generate a list of countries in the Americas to sort into North and South America 
```{r}
americas <- table(gapminder$country, gapminder$continent)[,2]
sort(americas, decreasing=T)
gapminder$icontinent <- ifelse(gapminder$country=="Argentina", "South America", ifelse(gapminder$country=="Bolivia", "South America", ifelse(gapminder$country=="Brazil", "South America", ifelse(gapminder$country=="Chile", "South America", gapminder$continent), ifelse(gapminder$country=="Colombia", "South America", ifelse(gapminder$country=="Ecuador", "South America", gapminder$continent), ifelse(gapminder$country=="El Salvador", "South America", ifelse(gapminder$country=="Guatemala", "South America", ifelse(gapminder$country=="Nicaragua", "South America", ifelse(gapminder$country=="Panama", "South America",  ifelse(gapminder$country=="Paraguay", "South America",ifelse(gapminder$country=="Peru", "South America",  ifelse(gapminder$country=="Trinidad and Tobago", "South America",  ifelse(gapminder$country=="Uruguay", "South America", ifelse(gapminder$country=="Venezuela", "South America", gapminder$continent)))))))))))))


gapminder$icontinent <- ifelse(gapminder$country=="Canada", "North America", gapminder$continent)
gapminder$icontinent <- ifelse(gapminder$country=="Costa Rica", "North America", gapminder$continent)
gapminder$icontinent <- ifelse(gapminder$country=="Cuba", "North America", gapminder$continent)
gapminder$icontinent <- ifelse(gapminder$country=="Dominican Republic", "North America", gapminder$continent)
gapminder$icontinent <- ifelse(gapminder$country=="Haiti", "North America", gapminder$continent)
gapminder$icontinent <- ifelse(gapminder$country=="Honduras", "North America", gapminder$continent)
gapminder$icontinent <- ifelse(gapminder$country=="Jamaica", "North America", gapminder$continent)
gapminder$icontinent <- ifelse(gapminder$country=="Mexico", "North America", gapminder$continent)
gapminder$icontinent <- ifelse(gapminder$country=="Puerto Rico", "North America", gapminder$continent)
gapminder$icontinent <- ifelse(gapminder$country=="United States", "North America", gapminder$continent)
```

### gap in life expectancy between countries on different continents narrowed over time 
```{r}
for (i in continents) {
  print(paste0("Continent: ", i))
  for (j in years) {
    mean_le <- mean(gapminder$log_lifeExp[gapminder$continent==i & gapminder$year == j], na.rm=T)
    mean_gdp <- mean(gapminder$vec_log_gdpPercap[gapminder$continent==i & gapminder$year==j], na.rm=T)
  print(paste0(j, " -> Mean Life Expectancy: ", mean_le, "; Mean GDP: ", mean_gdp))
  }
  }
```
```
