---
title: "Exercises Day 2"
author: "Richard Paquin Morel, adapted from exercises by Christina Maimone"
date: "`r Sys.Date()`"
output: html_document
params:
    answers: FALSE
---


```{r, echo=FALSE, eval=TRUE}
answers<-params$answers
```

```{r global_options, echo = FALSE, include = FALSE}
knitr::opts_chunk$set(echo=answers, eval=answers,
                      warning = FALSE, message = FALSE,
                      cache = FALSE, tidy = FALSE)
```

## Load the data

Load the `gapminder` dataset.

```{asis}
gapminder <- read.csv("data/gapminder5.csv", sep = ",", stringsAsFactors=FALSE)
```

```{r}
gapminder <- read.csv(here::here("data/gapminder5.csv"), stringsAsFactors=FALSE)
```

## Convert variables from factors to characters
```{r}
gapminder$country <- as.character(gapminder$country)
gapminder$continent <- as.character(gapminder$continent)
str(gapminder)
```

## mean life expectancy by country write a loop instead of writing each country name 
```{r}
mean(gapminder$lifeExp[gapminder$country=="Afghanistan"])
mean(gapminder$lifeExp[gapminder$country=="Albania"])
```

## use loop to recover GDP for each country 
Create a vector of values that you want to repeat the function for 
```{r}
obs <- 1:nrow(gapminder)
```
Initialize the for loop with for (i in vector)
```{r}
for (i in obs) 
  {gapminder[i, "gdp"] <- gapminder[i, "pop"]*gapminder[i, "gdpPercap"]}
```

## use loop to recover natural log of the GDP per capita & log of the population for each country 
Create a vector of values that you want to repeat the function for 
```{r}
obs <- 1:nrow(gapminder)
```
Initialize the for loop with for (i in vector). i is a placeholder and is replaced by subsequent code
```{r}
for (i in obs) 
  {gapminder[i, "log_gdpPercap"] <- log(gapminder[i, "gdpPercap"])
  gapminder[i, "log_pop"] <- log(gapminder[i, "pop"])}
```
Could also accomplish this via vectorized function
```{r}
gapminder$vec_log_gdpPercap <- log(gapminder$gdpPercap)
all(gapminder$vec_log_gdpPercap == gapminder$log_gdpPercap)
```

## life expectancy over time - find mean life expectancy by year
```{r}
years <- unique(gapminder$year)
for (i in years) {
  mean_le <- mean(gapminder$lifeExp[gapminder$year ==i], na.rm=T)
  print(paste0(i, ": ", mean_le))
}
```

## life expectancy over continent - which continent has highest mean life expectancy
```{r}
continents <- unique(gapminder$continent)
for (i in continents) {
  mean_lec <- mean(gapminder$lifeExp[gapminder$continent ==i], na.rm=T)
  print(paste0(i, ": ", mean_lec))
}
```
Could also accomplish this with aggregate()
```{r}
aggregate(gapminder$lifeExp ~ gapminder$continent, FUN=mean)
aggregate(lifeExp ~ continent, data=gapminder, FUN=mean)
```

## life expectancy for each continent for each year requires a nested loop to iterate over continent first
```{r}
for (i in continents) {
  print(paste0("Continent: ", i))
  for (j in years) {
    mean_le <- mean(gapminder$lifeExp[gapminder$continent==i & gapminder$ year == j], na.rm=T)
  print(paste0(j, ": ", mean_le))
  }
  }
```

## gap in life expectancy between countries on different continents narrowed over time 
```{r}
for (i in continents) {
  print(paste0("Continent: ", i))
  for (j in years) {
    sd_le <- sd(gapminder$lifeExp[gapminder$continent==i & gapminder$ year == j], na.rm=T)
  print(paste0(j, ": ", sd_le))
  }
  }
```

## instead of loops, apply 
select three variables 
```{r}
vars <- gapminder[, c("lifeExp", "pop", "gdpPercap")]
apply(vars, 2, mean)

for (i in vars) {
  print(mean(i))
}
```

## mean of every variable
```{r}
lapply(gapminder, mean)
sapply(gapminder, mean)
```

## anonymous functions in apply
```{r}
sapply(years, function(x) mean(gapminder$lifeExp[gapminder$year==x]))
```

## while loops, defining an iterator that will increase for each loop, ie every five years
```{r}
i <- 1952
while (i < 1987) {
  sd_le <- sd(gapminder$lifeExp[gapminder$year==i])
  print(paste0(i, ": ", sd_le))
  i <- i + 5
}
```

## while loop finding standard deviation for life expectancy for every 5 yearsbetween 1987 and 2002 
```{r}
i <- 1987
while (i <= 2002) {
  sd_le <- sd(gapminder$lifeExp[gapminder$year==i])
  print(paste0(i, ": ", sd_le))
  i <- i +5
}
```


## If Statement

## continents with mean life expectancy greater than 70 years
```{r}
threshold <- 70
for (i in unique(gapminder$continent)) {
  tmp <- mean(gapminder$lifeExp[gapminder$continent==i])
  if (tmp < threshold) {
    print(paste("Mean Life Expectancy in", i, "is less than", threshold))} 
    else {print(paste("Mean Life Expectancy in", i, "is greater than", threshold))}
  }
```

## report mean population for years greater than or equal to 1987 
```{r}
threshold <- 1987
for (i in unique(gapminder$year)) {
  tmp <- mean(gapminder$lifeExp[gapminder$year==i])
  if (tmp < threshold) {
    print(paste("Mean Life Expectancy in", i, "is less than", threshold))} 
    else {print(paste("Mean Life Expectancy in", i, "is greater than", threshold))}
  }
```

Use an if() statement to print a suitable message reporting whether there are any records from 2002 in the gapminder dataset. 

Hint: use the `any` function.

```{r}
year<-2002
if(any(gapminder$year == year)){
   print(paste("Record(s) for the year",year,"found."))
} else {
  print(paste("No records for year",year))
}
```

Now do the same for 2012.
```{r}
year<-2012
if(any(gapminder$year == year)){
   print(paste("Record(s) for the year",year,"found."))
} else {
  print(paste("No records for year",year))
}
```


## Loop and If Statements

Write a script that finds the mean life expectancy by country for countries whose population is below the mean for the dataset

```{r}
overall_mean <- mean(gapminder$pop)

for (i in unique(gapminder$country)) {
   country_mean <- mean(gapminder$pop[gapminder$country==i])
   
   if (country_mean < overall_mean) {
     mean_le <- mean(gapminder$lifeExp[gapminder$country==i])
     print(paste("Mean Life Expectancy in", i, "is", mean_le))
   } 
} # end for loop
```

Write a script that loops through the `gapminder` data by continent and prints out whether the mean life expectancy is smaller than 50, between 50 and 70, or greater than 70.

```{r}
lower_threshold <- 50
upper_threshold <- 70
 
for (i in unique(gapminder$continent)){
    tmp <- mean(gapminder$lifeExp[gapminder$continent==i])
    
    if (tmp < lower_threshold){
        print(paste("Average Life Expectancy in", i, "is less than", lower_threshold))
    }
    else if (tmp > lower_threshold & tmp < upper_threshold){
        print(paste("Average Life Expectancy in", i, "is between", lower_threshold, "and", upper_threshold))
    }
    else {
        print(paste("Average Life Expectancy in", i, "is greater than", upper_threshold))
    }
   
}
```


## Exercise: Write Functions

Create a simple funciton that prints the mean and standard deviation for life expectancy for a given country in the gapminder dataset

```{r}
report_mean_sd <-
  function(df, variable, country) {
    var <- df[[variable]][df$country == country]
    m_le <- mean(var)
    sd_le <- sd(var)
    cat("Country:", country,
        "\nMean Life Expectancy:", m_le,
        "\nSD Life Expectancy:", sd_le)
  }
report_mean_sd(gapminder, "lifeExp", "Bulgaria")
```

Create a funciton that reports the mean, median, minimum, and maximum for life expectancy for a continent in gapminder
```{r}
report_mean_sd <-
  function(df, variable, continent) {
    var <- df[[variable]][df$continent == continent]
    m_le <- mean(var)
    med_le <- median(var)
    min_le <- min(var)
    max_le <- max(var)
    cat("Country:", continent,
        "\nMean Life Expectancy:", m_le,
        "\nMedian Life Expectancy:", med_le,
        "\nMinimum Life Expectancy:", min_le,
        "\nMaximum Life Expectancy:", max_le)
  }
report_mean_sd(gapminder, "lifeExp", "Asia")
```

Create a function that given a data frame will print the name of each column and the class of data it contains. Use the gapminder dataset.  Hint: Use `mode()` or `class()` to get the class of the data in each column. Remember that `names()` or `colnames()` returns the name of the columns in a dataset.

Note: Some of these were taken or modified from https://www.r-bloggers.com/functions-exercises/

```{r}
data_frame_info <- function(df) {
  cols <- names(df)
  for (i in cols) {
    print(paste0(i, ": ", mode(df[, i])))
  }
}
data_frame_info(gapminder)
```

Create a function that given a vector will print the mean and the standard deviation of a **vector**, it will optionally also print the median. Hint: include an argument that takes a boolean (`TRUE`/`FALSE`) operator and then include an `if` statement.

```{r}
vector_info <- function(x, include_median=FALSE) {
  print(paste("Mean:", mean(x)))
  print(paste("Standard Deviation:", sd(x)))
  if (include_median) {
    print(paste("Median:", median(x)))
  }
}

le <- gapminder$lifeExp
vector_info(le, include_median = F)
vector_info(le, include_median = T)
```

## Analyzing the relationship 

Use what you've learned so far to answer the following questions using the `gapminder` dataset. Be sure to include some visualizations!

1. What is the relationship between GDP per capita and life expectancy? Does this relationship change over time? (Hint: Use the natural log of both variables.)
```{r}
gapminder$vec_log_gdpPercap <- log(gapminder$gdpPercap)
gapminder$log_lifeExp <- log(gapminder$lifeExp)
smoothScatter(gapminder$log_lifeExp, gapminder$vec_log_gdpPercap)
plot(gapminder$log_lifeExp, gapminder$vec_log_gdpPercap)
cor(gapminder$log_lifeExp, gapminder$vec_log_gdpPercap)
```

2. Does the relationship between GDP per capita and life expectacy vary by continent? Make sure you divide the Americas into North and South America.